---
title: "Lab 6 ESS-330"
format: html
editor: visual
---

Here I downloaded and loaded in all the necessary packages.

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = normalizePath(".."))
library(tidyverse)
library(tidymodels)
library(powerjoin)
library(glue)
library(vip)
library(baguette)
```
Now I have to download the CAMELS dataset and documentation PDF
```{r}
root  <- 'https://gdex.ucar.edu/dataset/camels/file'
```

```{r}
download.file(
  "https://gdex.ucar.edu/dataset/camels/file/camels_attributes_v2.0.pdf",
  destfile = here::here("data","camels_attributes_v2.0.pdf"),
  mode     = "wb"
)


```

After downloading the data, I had to create remote and local file paths for six CAMELS attribute text files, then I downloaded each one into my data folder. Then I read all the downloaded files into R as tibbles and performs a full outer join on gauge_id to combine them into a single dataset.
```{r}
types <- c("clim", "geol", "soil", "topo", "vege", "hydro")

remote_files  <- glue('{root}/camels_{types}.txt')

local_files   <- glue('data/camels_{types}.txt')

walk2(remote_files, local_files, download.file, quiet = TRUE)

camels <- map(local_files, read_delim, show_col_types = FALSE) 

camels <- power_full_join(camels ,by = 'gauge_id')
```

Question 1: zero_q_freq represents the frequency of days with Q = 0 mm/day. 

Here is the map that was provided to us in the lab instructions.
```{r}
library(ggthemes)
ggplot(data = camels, aes(x = gauge_lon, y = gauge_lat)) +
  borders("state", colour = "gray50") +
  geom_point(aes(color = q_mean)) +
  scale_color_gradient(low = "pink", high = "dodgerblue") +
  ggthemes::theme_map()
```

Question #2:
Here are my 2 maps
```{r}

library(patchwork)

# Map 1: colored by aridity
map_aridity <- ggplot(camels, aes(x = gauge_lon, y = gauge_lat)) +
  borders("state", colour = "gray50") +
  geom_point(aes(color = aridity), size = 1.5) +
  scale_color_viridis_c(option = "magma", name = "Aridity\n(index)") +
  labs(
    title    = "CAMELS Sites: Aridity",
    x        = "Longitude",
    y        = "Latitude"
  ) +
  theme_map() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 14, face = "bold")
  )

# Map 2: colored by mean precipitation
map_pmean <- ggplot(camels, aes(x = gauge_lon, y = gauge_lat)) +
  borders("state", colour = "gray50") +
  geom_point(aes(color = p_mean), size = 1.5) +
  scale_color_viridis_c(option = "plasma", name = expression(bar(P)~"(mm)")) +
  labs(
    title    = "CAMELS Sites: Mean Precipitation",
    x        = "Longitude",
    y        = "Latitude"
  ) +
  theme_map() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 14, face = "bold")
  )

# Combine side-by-side
map_aridity + map_pmean + 
  plot_layout(ncol = 2, guides = "collect") & 
  theme(legend.position = "bottom")

```

